### 2024-03-05 tex/dill.tex
[ correction ] first pass over chapter 3

I list the complete rules of DILL in \cref{fig:dill}.
Given \cref{def:DILL-empty,def:DILL-split,def:DILL-var}, the rules for
multiplicative and additive connectives look not too dissimilar to those in
\cref{fig:mall}.
However, their treatment of intuitionistic assumptions (contraction and
weakening automatically, as in non-substructural calculi) is new.
Meanwhile, the rules for the $\oc$-modality are the same as those for
the $\Box$-modality given in \cref{fig:PD}, except for the implicit structural
rules applying to linear assumptions.
For example, where the \TirName{$\Box$-I} rule discards true assumptions, the
\TirName{$\oc$-I} rule cannot discard linear assumptions, but instead requires
that there are no linear assumptions in the context when the rule is applied, so
that no linear assumptions can be used in the subproof.

\begin{figure}
  \begin{align*}
    A, B, C &\Coloneqq X \mid I \mid A \otimes B \mid A \multimap B
              \mid 0 \mid A \oplus B \mid \top \mid A \with B \mid \oc A \\
    \Gamma, \Delta, \Theta &\Coloneqq {\cdot} \mid \Gamma, A~\mathit{lin}
                             \mid \Gamma, A~\mathit{int} \\
    \mathcal S &\Coloneqq \Gamma \vdash A~\mathit{lin}
  \end{align*}
  \begin{mathpar}
    \ebrule{%
      \hypo{\Gamma \ni A}
      \infer1[Var]{\Gamma \vdash A~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash I~\mathit{lin}}
      \hypo{\Delta \vdash C~\mathit{lin}}
      \infer2[$I$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma~\mathit{int}}
      \infer1[$I$-I]{\Gamma \vdash I~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A \otimes B~\mathit{lin}}
      \hypo{\Delta, A~\mathit{lin}, B~\mathit{lin} \vdash C~\mathit{lin}}
      \infer2[$I$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A~\mathit{lin}}
      \hypo{\Delta \vdash B~\mathit{lin}}
      \infer2[$\otimes$-I]{\Gamma + \Delta \vdash A \otimes B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A \multimap B~\mathit{lin}}
      \hypo{\Delta \vdash A~\mathit{lin}}
      \infer2[$\multimap$-E]{\Gamma + \Delta \vdash B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma, A~\mathit{lin} \vdash B~\mathit{lin}}
      \infer1[$\multimap$-I]{\Gamma \vdash A \multimap B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash 0~\mathit{lin}}
      \infer1[$0$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \text{(no \TirName{$0$-I})}
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A \oplus B~\mathit{lin}}
      \hypo{\Delta, A~\mathit{lin} \vdash C~\mathit{lin}}
      \hypo{\Delta, B~\mathit{lin} \vdash C~\mathit{lin}}
      \infer3[$\oplus$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A_i~\mathit{lin}}
      \infer1[$\oplus$-I$_i$]{\Gamma \vdash A_0 \oplus A_1~\mathit{lin}}
    }
    \and
    \text{(no \TirName{$\top$-E})}
    \and
    \ebrule{%
      \infer0[$\top$-I]{\Gamma \vdash \top~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A_0 \with A_1~\mathit{lin}}
      \infer1[$\with$-E$_i$]{\Gamma \vdash A_i~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A~\mathit{lin}}
      \hypo{\Gamma \vdash B~\mathit{lin}}
      \infer2[$\with$-I]{\Gamma \vdash A \with B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash \oc A~\mathit{lin}}
      \hypo{\Delta, A~\mathit{int} \vdash \oc C~\mathit{lin}}
      \infer2[$\oc$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A~\mathit{lin}}
      \hypo{\Gamma~\mathit{int}}
      \infer2[$\oc$-I]{\Gamma \vdash \oc A~\mathit{lin}}
    }
  \end{mathpar}
  \caption{Dual Intuitionistic Linear Logic}
  \label{fig:dill}
\end{figure}

@@

Finally, notice that the rules given in \cref{fig:dill} characterise $\oc$ up to
logical equivalence.
The main feature ensuring this property is that each logical rule contains
exactly one occurrence of a logical connective, meaning that each logical rule
is only defining that connective in that place.
For the full characterisation result, we also require local soundness and
completeness of $\oc$, as discussed by \citet{judgmental} in the modal setting.

>>>>

I list the complete rules of DILL in \cref{fig:dill}.
Given \cref{def:DILL-empty,def:DILL-split,def:DILL-var}, the rules for
multiplicative and additive connectives look not too dissimilar to those in
\cref{fig:mall}.
However, their treatment of intuitionistic assumptions (contraction and
weakening automatically, as in non-substructural calculi) is new.
Meanwhile, the rules for the $\oc$-modality are the same as \TirName{$\Box$-I}
and \TirName{$\Box$-E} from \cref{sec:modal}, except for the implicit structural
rules applying to linear assumptions.
For example, where the \TirName{$\Box$-I} rule discards true assumptions, the
\TirName{$\oc$-I} rule cannot discard linear assumptions, but instead requires
that there are no linear assumptions in the context when the rule is applied, so
that no linear assumptions can be used in the subproof.

\begin{figure}
  \begin{align*}
    A, B, C &\Coloneqq X \mid I \mid A \otimes B \mid A \multimap B
              \mid 0 \mid A \oplus B \mid \top \mid A \with B \mid \oc A \\
    \Gamma, \Delta, \Theta &\Coloneqq {\cdot} \mid \Gamma, A~\mathit{lin}
                             \mid \Gamma, A~\mathit{int} \\
    \mathcal S &\Coloneqq \Gamma \vdash A~\mathit{lin}
  \end{align*}
  \begin{mathpar}
    \ebrule{%
      \hypo{\Gamma \ni A}
      \infer1[Var]{\Gamma \vdash A~\mathit{lin}}
    }
    \\
    \ebrule{%
      \hypo{\Gamma \vdash I~\mathit{lin}}
      \hypo{\Delta \vdash C~\mathit{lin}}
      \infer2[$I$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma~\mathit{int}}
      \infer1[$I$-I]{\Gamma \vdash I~\mathit{lin}}
    }
    \\
    \ebrule{%
      \hypo{\Gamma \vdash A \otimes B~\mathit{lin}}
      \hypo{\Delta, A~\mathit{lin}, B~\mathit{lin} \vdash C~\mathit{lin}}
      \infer2[$I$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A~\mathit{lin}}
      \hypo{\Delta \vdash B~\mathit{lin}}
      \infer2[$\otimes$-I]{\Gamma + \Delta \vdash A \otimes B~\mathit{lin}}
    }
    \\
    \ebrule{%
      \hypo{\Gamma \vdash A \multimap B~\mathit{lin}}
      \hypo{\Delta \vdash A~\mathit{lin}}
      \infer2[$\multimap$-E]{\Gamma + \Delta \vdash B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma, A~\mathit{lin} \vdash B~\mathit{lin}}
      \infer1[$\multimap$-I]{\Gamma \vdash A \multimap B~\mathit{lin}}
    }
    \\
    \ebrule{%
      \hypo{\Gamma \vdash 0~\mathit{lin}}
      \infer1[$0$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \text{(no \TirName{$0$-I})}
    \\
    \ebrule{%
      \hypo{\Gamma \vdash A \oplus B~\mathit{lin}}
      \hypo{\Delta, A~\mathit{lin} \vdash C~\mathit{lin}}
      \hypo{\Delta, B~\mathit{lin} \vdash C~\mathit{lin}}
      \infer3[$\oplus$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A_i~\mathit{lin}}
      \infer1[$\oplus$-I$_i$]{\Gamma \vdash A_0 \oplus A_1~\mathit{lin}}
    }
    \\
    \text{(no \TirName{$\top$-E})}
    \and
    \ebrule{%
      \infer0[$\top$-I]{\Gamma \vdash \top~\mathit{lin}}
    }
    \\
    \ebrule{%
      \hypo{\Gamma \vdash A_0 \with A_1~\mathit{lin}}
      \infer1[$\with$-E$_i$]{\Gamma \vdash A_i~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A~\mathit{lin}}
      \hypo{\Gamma \vdash B~\mathit{lin}}
      \infer2[$\with$-I]{\Gamma \vdash A \with B~\mathit{lin}}
    }
    \\
    \ebrule{%
      \hypo{\Gamma \vdash \oc A~\mathit{lin}}
      \hypo{\Delta, A~\mathit{int} \vdash \oc C~\mathit{lin}}
      \infer2[$\oc$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A~\mathit{lin}}
      \hypo{\Gamma~\mathit{int}}
      \infer2[$\oc$-I]{\Gamma \vdash \oc A~\mathit{lin}}
    }
  \end{mathpar}
  \caption{Dual Intuitionistic Linear Logic}
  \label{fig:dill}
\end{figure}

@@

Notice that the rules given in \cref{fig:dill} characterise $\oc$ up to
logical equivalence.
The main feature ensuring this property is that each logical rule contains
exactly one occurrence of a logical connective, meaning that each logical rule
is only defining that connective in that place.
For the full characterisation result, we also require local soundness and
completeness of $\oc$, as discussed by \citet{judgmental} in the modal setting.

As a concluding note on DILL, and as motivation for \cref{sec:semirings}, I
consider the extension of DILL to type dependency.
Such an extension has been studied by \citet{Vakar14}, and a similar extension
to linear/non-linear logic~\citep{Benton94} has been studied by \citet{KPB15}.
While these systems are well behaved on their own terms, both conspicuously omit
dependency on linear values.
This limitation means that one is not able to reason inside the language about
linear values --- for example, we cannot state equations between linear values.
The limitation arises from the fact that, in DILL and most other traditional
presentations of linear logic, linear variables go out of scope in any subterms
in which they are not used.
As noted by \citet{McBride16}, appearance in types should not be considered
\emph{use}.
According to DILL, unused linear variables are not in scope, and thus linear
variables are not in scope for types, and thus cannot appear in types.
\Citet{McBride16} develops a system, refined by \citet{Atkey18}, in which unused
linear variables remain in scope, and thus can appear in types.
It is this system, based on semiring annotations on variables, which is
most closely related to the system I present in \cref{sec:semirings}.

### 2024-03-05 tex/modal.tex
[ correction ] first pass over chapter 3

\begin{proposition}\label{thm:PT-NKT4}
  In NJ + \TirName{Promotion} + \TirName{T}, we can derive the original axioms
  and rule of S4: \TirName{N}, \TirName{K}, \TirName{T}, and \TirName{4}
\end{proposition}
\begin{proof}
  The rule \TirName{N} and axiom \TirName{T} are immediate.
  For axioms \TirName{K} and \TirName{4}, we have the following derivations,
  where \TirName{T$'$} abbreviates the obvious combination of \TirName{T} and
  \TirName{$\to$-E}.
  \begin{mathpar}
    \ebrule{%
      \infer0[T$'$]{\Box\plr{A \to B} \vdash A \to B}
      \infer0[T$'$]{\Box A \vdash A}
      \infer2[$\to$-E]{\Box\plr{A \to B}, \Box A \vdash B}
      \infer1[Promotion]{\Box\plr{A \to B}, \Box A \vdash \Box B}
      \infer1[$\to$-I]{\Box\plr{A \to B} \vdash \Box A \to \Box B}
      \infer1[$\to$-I]{\vdash \Box\plr{A \to B} \to \Box A \to \Box B}
    }
    \and
    \ebrule{%
      \infer0[Var]{\Box A \vdash \Box A}
      \infer1[Promotion]{\Box A \vdash \Box\Box A}
      \infer1[$\to$-I]{\vdash \Box A \to \Box\Box A}
    }
  \end{mathpar}
\end{proof}

\Cref{thm:NKT4-P,thm:PT-NKT4} suggest that a \TirName{Promotion}-like rule to
introduce the $\Box$ operator and a \TirName{T}-like rule to eliminate the
$\Box$ operator may be a possible approach to capture S4 in a natural deduction
style.
However, we do not want to add such a rule to our system because it is not
stable under substitution.
For example, let
$\sigma : \Box A \wedge \Box B \env\vdash \Box A, \Box B$ be the substitution
whose two components are given by left projection and right projection,
respectively.
The substitution principle says that we should be able to turn any derivation of
$\Box A, \Box B \vdash \Box C$ into a derivation of
$\Box A \wedge \Box B \vdash \Box C$.
However, if we concluded $\Box A, \Box B \vdash \Box C$ by \TirName{Promotion},
then it becomes unclear how to conclude $\Box A \wedge \Box B \vdash \Box C$.
In particular, \TirName{Promotion} does not apply because the formula
$\Box A \wedge \Box B$ is headed by $\wedge$, rather than $\Box$.

@@

The three new rules are listed in \cref{fig:PD}.
Let $\Gamma_v$ be the context $\Gamma$ but with all true hypotheses removed,
leaving only valid hypotheses.
Because I am working with a De Bruijn index-style presentation, weakening needs
to be admissible.
Therefore, the \TirName{$\Box$-I} rule needs to remove its true hypotheses,
rather than only applying in a context in which there are no true hypotheses.

\begin{figure}
  \begin{mathpar}
    \ebrule{%
      \hypo{\Gamma_v \vdash A~\mathit{true}}
      \infer1[$\Box$-I]{\Gamma \vdash \Box A~\mathit{true}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash \Box A~\mathit{true}}
      \hypo{\Gamma, A~\mathit{valid} \vdash B~\mathit{true}}
      \infer2[$\Box$-E]{\Gamma \vdash B~\mathit{true}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \ni \plr{A~\mathit{valid}}}
      \infer1[vVar]{\Gamma \vdash A~\mathit{true}}
    }
  \end{mathpar}
  \caption{The new rules of the \citeauthor{judgmental} presentation of IS4.}
  \label{fig:PD}
\end{figure}

@@

\begin{lemma}[lookup]\label{thm:PD-lookup}
  If $\vDash$ respects renaming (i.e., we have a function
  $\mathrm{ren}^\vDash :
  \Gamma \env\ni \Delta \to \Delta \vDash A \to \Gamma \vDash A$
  for each $\Gamma$, $\Delta$, and $A$), then
  from an environment $\rho : \Gamma \env\vDash \Delta$ and a variable
  $x : \Delta \ni A$, we get a value $\rho(x) : \Gamma \vDash A$.
\end{lemma}
\begin{proof}
  I proceed by cases on whether $x$ is a true or valid variable.
  In the true case, the result is straightforward.
  In the valid case, \cref{def:PD-env} gives us a value in $\Gamma_v \vDash A$.
  We need a renaming of type $\Gamma \env\vDash \Gamma_v$ to get the desired
  value in $\Gamma \vDash A$.
  The renaming is the one that, conceptually, discards the true hypotheses.
\end{proof}

>>>>

\begin{proposition}\label{thm:PT-NKT4}
  In NJ + \TirName{Promotion} + \TirName{T}, we can derive the original axioms
  and rule of S4: \TirName{N}, \TirName{K}, \TirName{T}, and \TirName{4}.
\end{proposition}
\begin{proof}
  The rule \TirName{N} and axiom \TirName{T} are immediate.
  For axioms \TirName{K} and \TirName{4}, we have the following derivations,
  where \TirName{T$'$} abbreviates the obvious combination of \TirName{T} and
  \TirName{$\to$-E}.
  \begin{mathpar}
    \ebrule{%
      \infer0[Var]{\Box\plr{A \to B} \vdash \Box\plr{A \to B}}
      \infer1[T$'$]{\Box\plr{A \to B} \vdash A \to B}
      \infer0[Var]{\Box A \vdash \Box A}
      \infer1[T$'$]{\Box A \vdash A}
      \infer2[$\to$-E]{\Box\plr{A \to B}, \Box A \vdash B}
      \infer1[Promotion]{\Box\plr{A \to B}, \Box A \vdash \Box B}
      \infer1[$\to$-I]{\Box\plr{A \to B} \vdash \Box A \to \Box B}
      \infer1[$\to$-I]{\vdash \Box\plr{A \to B} \to \Box A \to \Box B}
    }
    \and
    \ebrule{%
      \infer0[Var]{\Box A \vdash \Box A}
      \infer1[Promotion]{\Box A \vdash \Box\Box A}
      \infer1[$\to$-I]{\vdash \Box A \to \Box\Box A}
    }
  \end{mathpar}
\end{proof}

\Cref{thm:NKT4-P,thm:PT-NKT4} suggest that a \TirName{Promotion}-like rule to
introduce the $\Box$ operator and a \TirName{T}-like rule to eliminate the
$\Box$ operator may be a possible approach to capture S4 in a natural deduction
style.
However, we do not want to add \TirName{Promotion} to our system because it is
not stable under substitution.
For example, let
$\sigma : \Box A \wedge \Box B \env\vdash \Box A, \Box B$ be the substitution
whose two components are given by left projection and right projection,
respectively.
The substitution principle says that we should be able to turn any derivation of
$\Box A, \Box B \vdash \Box C$ into a derivation of
$\Box A \wedge \Box B \vdash \Box C$.
Let $A \coloneqq B \to C$.
Then, we have the following derivation of $\Box A, \Box B \vdash \Box C$, where
\TirName{T$'$} is the same abbreviation as it was in the proof of
\cref{thm:PT-NKT4}.

\begin{displaymath}
  \begin{prooftree}
    \infer0[Var]{\Box\plr{B \to C}, \Box B \vdash \Box\plr{B \to C}}
    \infer1[T$'$]{\Box\plr{B \to C}, \Box B \vdash B \to C}
    \infer0[Var]{\Box\plr{B \to C}, \Box B \vdash \Box B}
    \infer1[T$'$]{\Box\plr{B \to C}, \Box B \vdash B}
    \infer2[$\to$-E]{\Box\plr{B \to C}, \Box B \vdash C}
    \infer1[Promotion]{\Box\plr{B \to C}, \Box B \vdash \Box C}
  \end{prooftree}
\end{displaymath}

However, it is unclear how to derive
$\Box\plr{B \to C} \wedge \Box B \vdash \Box C$ in NJ + \TirName{Promotion} +
\TirName{T}.
In particular, \TirName{Promotion} does not apply because the formula
$\Box\plr{B \to C} \wedge \Box B$ is headed by $\wedge$, rather than $\Box$.

@@

The three new rules are listed below.
Let $\Gamma_v$ be the context $\Gamma$ but with all true hypotheses removed,
leaving only valid hypotheses.
Because I am working with a De Bruijn index-style presentation, weakening needs
to be admissible.
Therefore, the \TirName{$\Box$-I} rule needs to remove its true hypotheses,
rather than only applying in a context in which there are no true hypotheses.

\vspace{-2.5em}
\begin{mathpar}
  \ebrule{%
    \hypo{\Gamma_v \vdash A~\mathit{true}}
    \infer1[$\Box$-I]{\Gamma \vdash \Box A~\mathit{true}}
  }
  \and
  \ebrule{%
    \hypo{\Gamma \vdash \Box A~\mathit{true}}
    \hypo{\Gamma, A~\mathit{valid} \vdash B~\mathit{true}}
    \infer2[$\Box$-E]{\Gamma \vdash B~\mathit{true}}
  }
  \and
  \ebrule{%
    \hypo{\Gamma \ni \plr{A~\mathit{valid}}}
    \infer1[vVar]{\Gamma \vdash A~\mathit{true}}
  }
\end{mathpar}

@@

\begin{lemma}[lookup]\label{thm:PD-lookup}
  If $\vDash$ respects renaming (i.e., we have a function
  \mbox{$\mathrm{ren}^\vDash :
  \Gamma \env\ni \Delta \to \Delta \vDash A \to \Gamma \vDash A$}
  for each $\Gamma$, $\Delta$, and $A$), then
  from an environment $\rho : \Gamma \env\vDash \Delta$ and a variable
  $x : \Delta \ni A$, we get a value $\rho(x) : \Gamma \vDash A$.
\end{lemma}
\begin{proof}
  I proceed by cases on whether $x$ is a true or valid variable.
  In the true case, the result is straightforward.
  In the valid case, \cref{def:PD-env} gives us a value in $\Gamma_v \vDash A$.
  We need a renaming of type $\Gamma \env\vDash \Gamma_v$ to get the desired
  value in $\Gamma \vDash A$.
  The renaming is the one that, conceptually, discards the true hypotheses.
\end{proof}
